---
title: "RHAPSODY WP3 Pre-Diabetes Analysis Plan"
subtitle: "(Version 0.7.4)"
author: "Mickaël Canouil"
date: '`r format(Sys.Date(), "%d %B %Y")`'
output:  pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Files description
At this stage, you were provided with the following files:

  * `readme.pdf` the file you are actually reading;
  * `RHAPSODY_WP3_PreDiab.Rmd`, a Rmarkdown script which perform phenotype QC and preliminary modelling in a reproducible way;
  * `RHAPSODY_WP3_PreDiab.html`, an example of the output.

# Run the analysis (from section 1 to 6)
## R Setup
Please use `r R.version.string` (if not possible, newer version "3.4.*" can be used).  
Be sure the needed packages are installed or copy/paste the following command.
```{r}
check_packages <- function(package) {
  if (!package %in% installed.packages()[, "Package"]) {
    install.packages(
			package, 
			repos = c("https://rhap-fdb01.vital-it.ch/repo/", "https://cran.rstudio.com/", "http://cran.obiba.org"),
			dependencies = TRUE
		)
  } else {}
  library(package = package, character.only = TRUE)
}

list_packages <- c(
  "parallel",
  "grid",
  "scales",
  "broom",
  "viridis",
  "readxl",
  "writexl",
  "cowplot",
  "knitr",
  "kableExtra",
  "lme4",
  "lmerTest",
  "Hmisc",
  "data.tree",
  # "RCurl", 
  # "rjson",
  "opal",
  # "datashieldclient",
  # "dsCDISC",
  "tidyverse"
)

invisible(sapply(list_packages, check_packages))
```

`r if (knitr:::is_latex_output()) {'\\clearpage'} else {}`

## Run the Rmarkdown script
Only use the following parameters (according to your local RHAPSODY node information) from the command line (left all other Rmarkdown parameters to default).
```{r}
# set the output directory or leave at is; 
# output will be generated where the Rmarkdown file is
wd <- './' 

# Analysis
rmarkdown::render(
	input = 'RHAPSODY_WP3_PreDiab.Rmd', 
	output_format = "html_document", 
	output_dir = wd,
	params = list(
		cohort_name = 'CohortName',
		author_name = 'Firstname LASTNAME',
		opal_server = 'http://localhost:8080',
		opal_login = 'administrator',
		opal_password = 'password',
		output_directory = wd,
		# Don't change the following at this stage
		vcf_input_directory = './vcfs',
		imputation_quality_tag = 'INFO', 
		output_code = FALSE, 
		vcftools_binary_path = '/vcftools/vcftools_latest/bin', 
		format_vcfs = FALSE,
		variants_analysis = FALSE
	),
	encoding = 'UTF-8'
)
```

`r if (knitr:::is_latex_output()) {'\\clearpage'} else {}`

## Report results/bugs

### If everything ran smoothly
A document `RHAPSODY_WP3_PreDiab.html` should be available within the directory you specified (`wd`).  
Please check if there is no obvious errors in it and report back to mickael.canouil@cnrs.fr

### If something went wrong
In that case, you can try to run the Rmarkdown interactively using either Rstudio software to execute each chunk one at a time or copy/paste each chunk into R.  
In both cases you will have to copy/paste all parameters before running R code chunks in `"params_debug"` chunk on line 272 (set `eval=TRUE` and `include=TRUE`).

````markdown
`r ''````{r params_debug, eval = TRUE, include = TRUE}
params <- list(
	cohort_name = 'CohortName',
	author_name = 'Firstname LASTNAME',
	opal_server = 'http://localhost:8080',
	opal_login = 'administrator',
	opal_password = 'password',
	output_directory = wd,
	# Don't change the following at this stage
	vcf_input_directory = './vcfs',
	imputation_quality_tag = 'INFO',
	output_code = FALSE,
	vcftools_binary_path = '/vcftools/vcftools_latest/bin', 
	format_vcfs = FALSE,
	variants_analysis = FALSE
)
```
````

Once you identified where the error occurs, send the following information to mickael.canouil@cnrs.fr:

  * error messages (Screenshot or copy/paste it)
  * R code chunk where the error occurs
  * if possible, the result of `str(object)`, where object is the R object from which the error comes from (screenshot is fine)


